Accelerated Inference for Partially Observed Stochastic Processes via Automatic Differentiation

### 2024-03-14 mop/project.tex
Updates from Overleaf

\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}
\author{Kevin Tan, Giles J. Hooker, Edward L. Ionides}
\date{}

@@

\begin{document}
\maketitle
\begin{abstract}

Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods. Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques due to (1) the requirement for simulation-based inference that does not require access to the system's transition probabilities and (2) the issue that widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters. We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff. Further, we develop optimization algorithms suited to the Monte Carlo properties of the derivative estimate, including a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation. Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from iterated filtering can beat current state-of-the-art methods on a challenging scientific benchmark problem.
\end{abstract}

\section{Introduction}

@@

This is the backbone behind the only simulation-based full-information maximum likelihood method available in the literature thus far, the improved iterated filtering algorithm of \citet{ionides15}. \citet{ionides15} utilize the bootstrap filter for perform maximum likelihood estimation with an iterated perturbed Bayes map, successfully tackling challenging problems in epidemiology that available Bayesian methodology fails to solve.

Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small. For example, though in practice the algorithm of \citet{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood. \citet{Ionides_mcap} attempt to address the third problem, albeit with a computationally-expensive approach.

\subsection{Automatic Differentiation for Particle Filters}

@@

\citet{blei2018vsmc} and \citet{jon2018diffpf} drop high-variance resampling terms from the gradient, leading to bias that does not vanish asymptotically according to \citet{corenflos21}. Though the approach of \citet{corenflos21} is promising, it is computationally expensive. \citet{doucet2022particlebased} provide another promising fixed-lag smoothing-based approach that is similar in spirit to our method (MOP-$\alpha$), but requires either (1) access to the process model transition densities or (2) the special case where the transition density factors into a policy that selects actions (in the reinforcement learning sense) and a deterministic model that maps states and actions to next-timestep states.

Finally, \citet{scibior2021dpf} demonstrate that the estimators derived in \citet{poyiadjis11} can be attained with AD, albeit with a simple tweak to the particle filter that does not modify its likelihood estimates. Though their method is not fully compatible with simulation-based inference, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.

\subsection{Our Contributions}

@@

We show these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties. To cap off the theoretical results, we also derive a linear convergence rate for gradient methods on the particle filter in the presence of strong convexity, providing support for numerical optimization with AD for the particle filter.

In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering. While this requires a differentiable simulator compatible with the reparametrization trick (as previous literature such as \citet{corenflos21} also require), future work will deal with workarounds inspired by our framework that use likelihood ratios of transition densities in place of this. Promising numerical results indicate that IFAD beats IF2 (and by the transitive property, IF1, the Liu-West filter, and particle MCMC) on a challenging scientific benchmark problem.

While the Monte-Carlo Adjusted Profile (MCAP) from \citet{Ionides_mcap} provides a profile likelihood confidence interval that takes into account both Monte Carlo profile error and statistical uncertainty in the likelihood, it is computationally inefficient. \kevin{TODO} We provide a method for incorporating the MCAP methodology (with the poor man's likelihood profile) in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval. 
\kevin{Quantile regression for estimating the max/the envelope of the likelihood evaluations from AD, then fit MCAP/MQLE/MSLE?}

Finally, we make the first steps towards developing a Python counterpart to the popular \texttt{pomp} R package by \citet{king16, king2017pompmanual} that we call \texttt{pompy}. The software we provide includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation enabled by the \texttt{jax} Python package from \citet{jax2018github}.

\kevin{Terminology: Plug-and-play (clearly definable, but not well-known, define at start of paper and don't use in title/abstract), likelihood-free maximum likelihood (seemingly an oxymoron, rationale being there is no need to evaluate density, bayesians call likelihood densities, plus we only approximate the likelihood and don't really have it), simulation-based inference (loaded term)}

\subsection{Significance Statement}
Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements. Under the Markov assumption on system dynamics, previous work has provided methods of performing inference for these models. In particular, prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics. We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation. Our new method outperforms previous approaches on a challenging problem in epidemiology. This is implemented in a precursor to a Python counterpart to the popular \texttt{pomp} package by \citet{king16, king2017pompmanual} that includes welcome features for scientific computing such as GPU acceleration, parallel computing for single-program-multiple-data programs, just-in-time compilation, and automatic differentiation.

\kevin{TODO: (1) Fix beta smoothness to d(n) at no more than O(n), (2) add story to say that no one spotted that it could have been plug-and-play because of e.g. reasons why they would stumble}

\section{Problem Setup}

Consider an unobserved Markov process $\{X_t, t \geq t_0\}$, and observations $Y_1,...,Y_N$ at timesteps $t_1,..., t_N$. The process is parameterized by an unknown parameter $\theta \in \Theta$, where $X \in \gX, Y \in \gY,$ and finally $\gX, \gY, \Theta \subseteq \R^n$. By a similar decomposition to that in \citet{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
$$f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).$$

\paragraph{Seed-Fixing, Coupling, and the Method of Common Random Numbers, Coupling:}

In the rest of this paper, we will identify the seed with an element of the sample space. We do so because the seed determines the entire sequence of pseudo-random numbers as generated by a computer, to maintain an algorithmic and practical interpretation of our theory. When two algorithms share the same seed, this corresponds to the method of common random numbers.

\section{A Smooth Extension of the Particle Filter}

%\paragraph{Motivation:} 
%The gradient of the log-likelihood can be written as a vector $\nabla_\theta \ell(\theta)$ such that 
%\begin{equation}
    %\lim_{v \to 0} \frac{|\ell(\theta+v) - \ell(\theta) - v^T \nabla_\theta \ell(\theta)|}{\|v\|} = 0.
%\end{equation}

%We therefore need to evaluate the likelihood at a point $\theta$ and a nearby point $\theta + v$ to estimate the gradient with finite differences. But in the presence of Monte Carlo noise in evaluating the likelihood, this is difficult, unless (1) one ensures one uses the same seed for each evaluation in the neighborhood, and (2) chooses a step-size small enough that changing the the parameter does not change the resampling indices.

\paragraph{Intuition:} 
Our theoretical framework for automatic differentiation for the particle filter hinges on the following observation. Instead of seeking to differentiate $\ell(\theta)$ directly, we instead focus on differentiating through a suitable fixed-seed reweighting of a bootstrap filter. The intuition is as follows. Consider a particle filter likelihood estimate where the system (i.e. \textbf{both the state transitions and resampling}) evolves under a parameter $\phi \in \Theta$ and the seed, or sample space element, is fixed at $\omega \in \Omega.$ Conditional on $\omega$ and $\phi,$ one can reweight the likelihood estimate to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$

Differentiating this reweighted likelihood estimate with respect to $\theta$ then amounts to differentiating through the reweighting likelihood ratios, which is differentiable if the densities used for the reweighting are differentiable. We call this simulation-and-reweighting scheme an \textbf{off-policy} likelihood estimate, as it is similar in spirit to the off-policy methods found in the reinforcement learning literature.

The above intuition is formalized below.

\kevin{Method of Common Random Numbers, Coupling}

\subsection{Doubly Off-Policy Particle Filters}

\begin{defn}[Seed-Fixing]
\label{defn:seed-fixing}
Consider a probability space $(\Omega, \Sigma, \mu)$ governing the particles and observations. Noting that the particle filter is unbiased for the likelihood when computing the expectation of the normalizing factor under the posterior $f_{X_{0:N|Y_{1:N}}}\left(x_{0: N} \mid y_{1: N}; \theta\right)$, we can write the likelihood as
$$
\begin{aligned}
\mathcal{L}(\theta) & =\int_{\Omega} \mathcal{L}(\theta, \omega, J) d \mu(\omega) \\
& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right)\right) f_{X_{0:N}, Y_{1:N}}\left(x_{0: N}^P(\omega) , y_{1: N}^*, \theta\right) d \mu(\omega) \\
& =\int_{\Omega}\left(\prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J f_{Y_n|X_n}\left(y_n^* \mid x_{n, j}^P(\omega), \theta\right) f_{X_n|X_{n-1}}(x_{n,j}^P(\omega)|x_{n-1,j}^F(\omega), \theta)\right) f_{X_{0:N}| Y_{1:N}}\left(x_{n, j}^P(\omega) \mid y_n^*, \theta\right) d \mu(\omega)
\end{aligned}
$$
where $\mathcal{L}(\theta, \omega, J)$ is the Monte-Carlo estimate of the likelihood given the realization of $\mathrm{J}$ particles corresponding to $\omega \in \Omega$. For simplicity, we resample every timestep. We call $\omega$ a seed, and the deterministic quantity $\mathcal{L}(\theta, \omega, J)$ the seed-fixed estimate.
\end{defn}

Armed with this seed-fixed likelihood estimate, we modify a particle filter likelihood estimate where the system evolves under a parameter $\phi \in \Theta$ and seed $\omega \in \Omega$ to instead evaluate the likelihood at another (sufficiently nearby) $\theta \in \Theta.$

\begin{defn}[Doubly Off-Policy Particle Filter Likelihood]
\label{defn:doubly-off-policy}
Let $\lik(\theta, \phi, \omega, J)$ denote a Monte-Carlo estimate of the likelihood evaluated at $\theta \in \Theta$ with $J$ particles, but where the system evolves according to another $\phi \in \Theta$ and conditional on a fixed seed $\omega \in \Omega$. This can be computed with a suitable reweighting of the particles, which we provide the derivation for later. We then have
\begin{equation} \label{eq:loglik2}
\lik(\theta)=\E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega),
\end{equation}
where we define
\begin{align}
    \lik(\theta,\phi,\omega, J) = \prod_{n=1}^N \frac{1}{J} \sum_{j=1}^J &\left(f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi}(\omega), \phi) \cdot 
    f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}(\omega) | x_{n-1,j}^{F,\phi}, \phi) \right.\\
    &\cdot
    \left.
    \frac{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\theta)}{f_{Y_n|X_n}(y^*_n|x_{n,j}^{P,\phi}(\omega),\phi)}
    \cdot \frac{f_{X_n|X_{n-1}}(x_{n,j}^{P,\phi}(\omega)|x_{n-1,j}^{F,\phi}(\omega),\theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P,\phi}(\omega)|x_{n-1,j}^{F,\phi}(\omega),\phi)}\right).
\end{align}
\end{defn}

Lastly, we require the following regularity conditions.

\begin{aspt}[Continuity of the Likelihood] $\ell(\theta)$ proper is continuous in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$.
\end{aspt}

\begin{aspt}[Bounded Process Model]
\label{assump:bounded-process}
    There exist $\underbar{M}, \bar{M}$ such that $0 < \underbar{M} \leq f_{X_n|X_{n-1}}(x_n | x_{n-1}) \leq \bar{M} < \infty$. 
\end{aspt}

\begin{aspt}[Bounded Measurement Model]
\label{assump:bounded-measurement}
There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G} < \infty$ and $\infty > G' \geq G'(\theta)>0$ such that $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty < G'(\theta) \leq G'$ for all $1 \leq n \leq N, x_n \in \gX$ and $\theta \in \Theta.$
\end{aspt}

\begin{aspt}[Locally Bounded Derivatives]
\label{assump:local-bounded-derivative}
Let $\mathcal{M}$ be an open subset of $\Omega$. There exists some function $G(\theta)$ and constants $G=\sup _{\theta \in \mathcal{N}} G(\theta)<\infty$, $H=\sup _{\theta \in \mathcal{N}} H(\theta)<\infty$ such that
    $$
    \|\nabla_\theta \ell(\theta, \phi, \omega, J)\|_\infty < NG'(\theta) \leq G(\theta) \leq G<\infty, \;\;
    \|\nabla^2_\theta \ell(\theta, \phi, \omega, J)\|_\infty <N^2G'(\theta)^2 \leq G(\theta)^2 \leq G^2<\infty
    $$
    for every $\phi \in \mathcal{M}, \theta \in \mathcal{N}_J(\phi)$, and almost every $\omega \in \Omega$.
\end{aspt}

We need the following assumption to ensure that the weighted particle filter with off-policy resampling targets the posterior:
\begin{aspt}[Bounded And Differentiable Measurement Density Ratios]
    \label{assump:diff-meas}
    For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{Y_n|X_n}(y_n^*|x, \theta)}{f_{Y_n|X_n}(y_n^*|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x \in \gX$ and all $\theta\in\gN(\phi)$. 
\end{aspt}

In the following assumption, the first condition is sufficient for DOP, and the second for MOP:
\begin{aspt}[Bounded And Differentiable Process Model]
    \label{assump:diff-process}
    Either one of the following two conditions hold:
    \begin{enumerate}
        \item For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{X_n|X_{n-1}}(x'|x, \theta)}{f_{X_n|X_{n-1}}(x'|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x, x' \in \gX$ and all $\theta\in\gN(\phi)$, or
        \item There exists a differentiable simulator with more than two continuous derivatives for the process model. 
    \end{enumerate}
\end{aspt}

We want $\phi\in\Theta$ to be a point in parameter space around which $\lik(\theta,\phi,\omega, J)$ is differentiable with respect to $\theta$ and does not have excessive Monte~Carlo variance.

We require many of these assumptions to establish the following proposition, the proof of which is a straightforward consequence of what we assumed.

\begin{prop}[Smooth Neighborhood]
\label{prop:smooth-nbhd}
There exists an open neighborhood $\mathcal{N}(\phi)$ around any $\phi \in \Theta$ where for all $\theta \in \mathcal{N}(\phi)$ and almost every $\omega \in \Omega$, the Monte Carlo estimate of the likelihood at $\theta \in \mathcal{N}(\phi)$ with the system evolving according to $\phi$ conditional on $\omega$, $\lik(\theta,\phi,\omega, J)$, has two continuous derivatives in $\theta$.
\end{prop}

\begin{proof}
    For a fixed $\omega$, the resampling indices for $\lik(\theta,\phi,\omega, J)$ and $\lik(\phi,\phi,\omega, J)$ are the same as the resampling indices in both cases correspond to that yielded by $f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}(\omega);\phi)$, eliminating any discontinuities from resampling.

    By Assumptions \ref{assump:diff-meas} and \ref{assump:diff-process}, for small enough $\mathcal{N}(\phi)$ the density ratios are bounded. The likelihood only changes by a factor of the density ratios, which are bounded and possess two continuous derivatives in $\theta$ if either the process and measurement densities used in the calculation do, or if the measurement densities do and one has access to a differentiable simulator with two continuous derivatives in $\theta$. This gives us the desired result.

    As a remark, we note that the neighborhoods the sequence is differentiable in are independent of $J$, as the resampling does not change with $\theta$ -- it is dependent on $\phi$. 
\end{proof}

\subsubsection{Doubly Off-Policy Particle Filter}

Having the system evolve according to $\phi$ but evaluating the likelihood at $\theta$ requires us to reweight the conditional likelihood at each timestep by likelihood ratios corresponding to the process and measurement models. This corresponds to making two corrections in the particle weights at each timestep, by multiplying by 
\begin{equation}
    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
\end{equation}
as we illustrate in Appendix \ref{app:dop}. We use the notation $x_{n,j}^{P, \phi}$ to emphasize that the particles evolve according to $\phi$.

We claim that this yields a particle filter that targets the posterior and recovers the derivative estimate of \citet{poyiadjis11, scibior2021dpf} when $\theta$ is evaluated at $\phi$. The proof can be found in Appendix \ref{app:dop}.

\begin{prop}[Correctness of Doubly Off-Policy Particle Filter Corrections]
    \label{prop:dop-correctness}

    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$. Also let $\theta \in \gN(\phi)$. Then, the following corrections in the particle weights,
    \begin{equation}
    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
    \end{equation}
    yield a particle filter that is unbiased for the likelihood, even when $\theta \neq \phi$.

    In particular, when $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{poyiadjis11},
    \begin{equation}
        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
    \end{equation}
\end{prop}

\paragraph{Intuition:} Having the system evolve according to $\phi$ but evaluating the likelihood at $\theta$ requires us to reweight the conditional likelihood at each timestep by likelihood ratios corresponding to the process and measurement models. This corresponds to making two corrections in the particle weights at each timestep, one for the process model and one for the measurement model. The measurement correction cannot be avoided, but one can bypass the process correction -- with the reparametrization trick. We can do when we assume the simulator is differentiable with the reparametrization trick, as \cite{corenflos21} also assume. Doing so then yields an algorithm that only requires a simulator, and does not require the process densities.

\subsection{Measurement Off-Policy Particle Filters}

With the above procedure, we have the entire system evolve according to $\phi$, and we differentiate through the likelihood ratios, which are smooth conditional on the resampling being the same (which in turn happens when $\theta \approx \phi$). However, doing so requires access to likelihood ratios, and as such Algorithm \ref{alg:dop} is therefore not compatible with simulation-based inference.

There is a workaround. Consider the case where you have access to a differentiable simulator compatible with the reparametrization trick. If we have the state transitions evolve according to $\theta,$ but have the resampling according to $\phi,$ we can offload the derivative with respect to the process model to the reparametrization trick, as in \cite{corenflos21}. This corresponds to the MOP-$\alpha$ procedure in Algorithm \ref{alg:mop}.

We therefore make the following assumption in this section only:
\begin{aspt}[Differentiable Simulator]
    $\texttt{process}(\cdot | X_{t-1}; \theta)$ is differentiable with respect to $\theta,$ and is compatible with the reparametrization trick. 
\end{aspt}

Algorithm \ref{alg:mop} provides a weighted particle filter that targets the posterior under $\theta$, but when differentiated yields either the low-bias but high-variance gradient estimator of \cite{poyiadjis11}, the high-bias but low-variance estimator of \cite{blei2018vsmc}, or an intermediate, interpolating quantity that can optimize a bias-variance tradeoff. This is controlled by a discounting parameter $\alpha$. The intuition behind this procedure is as follows.

\paragraph{Intuition:} Broadly, if $\theta\neq\phi,$ one can first run a vanilla particle filter to recover values for the particles with the process model under $\phi.$ One can then run an accordingly reweighted particle filter with the process model under $\theta$, but resampling according to the measurement model under $\phi$, with the same indices as the original vanilla particle filter. This requires one to set the same random seed for both particle filters. This then yields a slightly-unconventionally weighted particle filter that targets the posterior, but has some very nice gradient properties; as we outline in the remainder of this section.

\paragraph{Special Case:} When $\theta = \phi$, one can simply set the particles under $\phi$ to be copies of the particles under $\theta$. Gradients with respect to $\theta$ now treat functions of the particles under $\phi$ as constants, corresponding to the stop-gradient trick in \cite{scibior2021dpf}. This therefore provides a theoretical framework for their stop-gradient trick, which we happily use in our numerical experiments for simplicity.

\kevin{Develop theory for mixing-time swapping? Detect when IF2 reaches a neighborhood of the MLE, by detecting whether it has reached a local sort of stationary distribution when the perturbations are lower bounded, e.g. the autocorrelation between the mean vectors to estimate the mixing time or Gelman-Rubin. In practice, just swap when IF2 converges according to the diagnostics. Theory doesn't care, you just need to in a basin of attraction that is strongly convex.}

\subsubsection{Discounting Weights}

Observe that at Algorithm \ref{alg:mop}, at each timestep the corrected filtering importance weights $w_{n,j}^{F,\theta}$ optionally propagate to the next timestep for us to maintain the correction. We introduce a discounting parameter $\alpha$ that affects the weights, so $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^{\alpha}$. Note that no matter what value $\alpha$ takes on, the system evolves according to $\phi$ when we perform the correction to evaluate the likelihood at $\theta$. The filter fully targets the posterior when $\theta=\phi$, or $\alpha=1$, as we show in Proposition \ref{prop:dop-correctness}. As previously mentioned, the value of $\alpha$ therefore informs a bias-variance tradeoff. We elaborate on this below.

\paragraph{Bias-Variance Tradeoff} This discounting serves the purpose of balancing a tradeoff between maintaining memory of each particle's ancestral trajectory (most extreme when $\alpha=1$), or considering only the single-step transition dynamics (when $\alpha=0$). This can be thought of as an \textbf{exponentially-weighted moving average}, where $\alpha$ controls the amount of discounting. We elaborate on two special cases below.

\kevin{This isn't a smoother, because we'd need future information.}

>>>>

\title{Accelerated Inference for Partially Observed Stochastic Processes via Automatic Differentiation}
\author{Kevin Tan}
\date{}

@@

\begin{document}

\maketitle

\begin{center}
    Under the supervision of Giles J. Hooker and Edward L. Ionides
\end{center}

\section{Background}
A partially-observed Markov process (POMP) is defined as follows. Consider an unobserved Markov process $\{X_t, t \geq t_0\}$, and observations $Y_1,...,Y_N$ at timesteps $t_1,..., t_N$. The process is parameterized by an unknown parameter $\theta \in \Theta$, where $X \in \gX, Y \in \gY,$ and finally $\gX, \gY, \Theta \subseteq \R^n$. By a similar decomposition to that in \citet{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
$$f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).$$

\begin{figure}[H]
\centering
\begin{tikzpicture}[scale=2]

\node[black] (X_0) at (-3,0) {$X_0$};
\node[black] (Xdot0) at (-2,0) {$...$};
\node[black] (X_{t-1}) at (-1,0) {$X_{t-1}$};
\node[black] (X_t) at (0,0) {$X_t$};
\node[black] (X_{t+1}) at (1,0) {$X_{t+1}$};
\node[black] (Xdot{t+1}) at (2,0) {$...$};
\node[black] (X_T) at (3,0) {$X_T$};
\node[black] (Y_0) at (-3,-1) {$Y_0$};
\node[black] (Ydot0) at (-2,-1) {$...$};
\node[black] (Y_{t-1}) at (-1,-1) {$Y_{t-1}$};
\node[black] (Y_t) at (-0,-1) {$Y_t$};
\node[black] (Y_{t+1}) at (1,-1) {$Y_{t+1}$};
\node[black] (Ydot{t+1}) at (2,-1) {$...$};
\node[black] (Y_T) at (3,-1) {$Y_T$};
\path[->, black] (X_0) edge (Xdot0);
\path[->, black] (Xdot0) edge (X_{t-1});
\path[->, black] (X_{t-1}) edge (X_t);
\path[->, black] (X_t) edge (X_{t+1});
\path[->, black] (X_{t+1}) edge (Xdot{t+1});
\path[->, black] (Xdot{t+1}) edge (X_T);
\path[->, black] (X_0) edge (Y_0);
\path[->, black] (X_{t-1}) edge (Y_{t-1});
\path[->, black] (X_t) edge (Y_t);
\path[->, black] (X_{t+1}) edge (Y_{t+1});
\path[->, black] (X_T) edge (Y_T);
\end{tikzpicture}
\caption{Directed acyclic graph (DAG) of a POMP model.}
\end{figure}

We seek to solve the filtering problem, that is, the problem of estimating
$$p(x_n|y_{1:n}),$$
or the posterior distribution of the state given the current and all previous observations. This can be obtained by Bayesian updating, namely that
\begin{enumerate}
    \item Given a belief on the state space $x_{n-1}$, simulate forward according to $p(\hat x_{n}|x_{n-1})$ to obtain a prediction distribution $\hat x_n$ for the state at time $n$, and
    \item Given a prediction $\hat{x}_n$ at time $n$, observe the measurement $y_n^*$ and perform a Bayesian update to the prediction according to $p(y_n^* | \hat x_{n-1})$ to obtain a belief on the state at time $n$, $x_n$. 
\end{enumerate}

The particle filter is simply the Monte Carlo approximation to this procedure, obtained by maintaining a fixed number $J$ of simulations that we call particles. This induces a mixture Dirac measure on the state space, where we call the belief approximation $x_{n,j}^F$ to the filtering distribution $p(x_n|y_{1:n})$, and the prediction approximation $x_{n,j}^P$ to the prediction distribution $p(x_{n}|y_{1:n-1})$. The intractable integrals involved in the Bayesian update are circumvented with sums over the Dirac measures.

At a high level, a particle filter keeps track of a set of particles $x_{n,j}$, sequentially approximating (1) the distribution of $X_{n}|y_{1:n-1}^*$ with a mixture Dirac measure $\frac{1}{J}\sum_{j=1}^J \tilde{w}_{n,j}\delta_{x_{n,j}^P}$ via simulation from the process model $p(x_n|x_{n-1})$ and (2) the distribution of $X_{n}|y_{1:n}^*$ with another mixture Dirac measure $\frac{1}{J}\sum_{j=1}^J w_{n,j}\delta_{x_{t,j}^P}$ via a reweighting of the particles in proportion to their likelihood.

The particle filter approximates the Bayes filter through these two sequences of mixture Dirac measures. The Bayes filter is the "optimal" filter that updates the prediction distribution $p(x_{t}^P|y_{1:t-1}^*)$ with the prediction formula from \cite{sbied_lec3}, 
\begin{equation}
    p(x_{n}^P|y_{1:n-1}^*) = \int p(x_{n-1}^F|y_{1:n-1}^*)p(x_t|x_{t-1}) dx_{n-1},
\end{equation}
and the filtering distribution $p(x_t^F|y_{1:t}^*)$ via Bayes' theorem \citep{sbied_lec3},
\begin{equation}
    p(x_n^F|y_{1:n}^*) = \frac{p(y_n^*|x_n)p(x_n|y_{1:n-1}^*)}{\int p(y_n^*|x_n)p(x_n|y_{1:n-1}^*) dx_n}.
\end{equation}

This therefore provides a simple, general, and computationally efficient solution, yielding an approximation to the filtering distribution and log-likelihood.

\section{Challenges}

@@

This is the backbone behind the only simulation-based full-information maximum likelihood method available in the literature thus far, the improved iterated filtering (IF2) algorithm of \citet{ionides15}. \citet{ionides15} utilize the bootstrap filter for perform maximum likelihood estimation with an iterated perturbed Bayes map, successfully tackling challenging problems in epidemiology that available Bayesian methodology fails to solve.

Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small. For example, though in practice the algorithm of \citet{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood, as we see below in Figure \ref{fig:if2fail}.

\begin{figure}[H]
    \centering
    \includegraphics[scale=0.5]{imgs/095/if2fail.png}
    \caption{Failure of iterated filtering on maximum likelihood estimation in the Dhaka cholera model of \cite{king08}. IF2 was run 100 times (with the 0th and 80th quantiles represented with the shaded area), but no run finds the MLE. The best run is about 7.5 log-likelihood units short. }
    \label{fig:if2fail}
\end{figure}

\section{A Potential Solution}

Given the success of stochastic gradient methods in machine learning, it is natural to consider the performance of the procedure of performing gradient descent on the log-likelihood estimate given by the particle filter. We seek to develop a method that performs better than IF2 on this problem, by running IF2 for a few iterations and taking advantage of its quick convergence to a neighborhood of the MLE, and then switching to stochastic gradient descent on the gradient estimate given by running automatic differentiation on the log-likelihood estimate from the particle filter.

\section{Automatic Differentiation for Particle Filters}

@@

In particular, only \cite{scibior21}, \cite{blei2018vsmc}, and \cite{corenflos21} have algorithms that are amenable to simulation-based inference without access to transition densities. Though the approach of \citet{corenflos21} is promising, it is computationally expensive, requiring runtime quadratic in the number of particles (unlike other approaches, which run in linear-time).

\cite{scibior21} show that the estimator of \cite{blei2018vsmc} corresponds to the estimate of the gradient given by differentiating through the vanilla particle filter. This is therefore a very natural estimator for the score. However, \citet{blei2018vsmc} drops high-variance resampling terms from the gradient, leading to bias that does not vanish asymptotically according to \citet{corenflos21}.

A linear-time consistent estimator of the score can be obtained via the estimator of \cite{poyiadjis11}, which \cite{scibior21} show can be obtained through a minor tweak on the particle filter that they call a "stop-gradient trick". However, this estimator has $O(N^2)$ variance, and the stop-gradient trick is not mathematically motivated.

Of the other methods that are not compatible with simulation-based inference, \citet{doucet2022particlebased} provide another promising fixed-lag smoothing-based approach, but requires either (1) access to the process model transition densities or (2) the special case where the transition density factors into a policy that selects actions (in the reinforcement learning sense) and a deterministic model that maps states and actions to next-timestep states.

\section{Our Contributions}

@@

We will show that these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties. This entails novel proofs showing a strong law of large numbers for triangular arrays under bounded particle weights and functionals, with the application of showing that a suitably reweighted particle filter still targets the posterior, even when resampling is carried out under some other arbitrary categorical distribution.

In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering. Promising preliminary numerical results indicate that this can beat current state-of-the-art methods on a challenging scientific benchmark problem.
